Aligning marine species range data to better serve science and conservation

### 2016-03-04 manuscript-main.Rmd
updating refs

---
title: "Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>"
output:
  word_document: default
  pdf_document:
    number_sections: yes
    toc: no
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
---

>>>>

---
title: "Aligning marine species range data to better serve science and conservation"
author: "Casey O'Hara^1^, Jamie Afflerbach^1^, Benjamin S. Halpern<sup>1, 2, 3</sup>"
output:
  html_document:
    highlight: haddock
    number_sections: yes
    theme: spacelab
    toc: no
  pdf_document:
    number_sections: yes
    toc: no
  word_document: default
---

### 2016-03-04 ms_1_abst_sig.Rmd
updating refs

Species distribution data provide the foundation for a wide range of ecological research and conservation science and management, yet most species ranges remain unknown, and existing range maps have little overlap. In the ocean, two global species distribution datasets, produced by AquaMaps and the International Union for Conservation of Nature (IUCN), dominate our understanding of marine species ranges throughout the world's oceans.  Together they represent `r format(nrow(spp_all), big.mark = ',')` species, mostly from AquaMaps, with only `r format(nrow(spp_map_pairs), big.mark = ',')` species overlapping.  Here we examine differences in predicted species ranges between the two datasets, propose mechanistic causes and potential solutions for such differences, and explore the implications of these differences for management and conservation decisions.  We find that IUCN maps often disregard bathymetry for depth-limited species, leading to predictions of species presence at unsuitable depths, and AquaMaps ranges for data-poor species often extrapolate presence far afield from known occurrences.  We illustrate the implications of these differences by repeating two recent applications - the Ocean Health Index biodiversity goal and a global analysis of gaps in coverage of marine protected areas - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on which dataset was used. ___Creating a single, highly accurate dataset of global marine species ranges will be difficult. Understanding the implications of dataset differences for conservation planning and decision-making is essential.___
<!--Understanding these issues points toward solutions for providers and users of these datasets.-->

>>>>

Species distribution data provide the foundation for a wide range of ecological research and conservation science and management, yet most species ranges remain unknown, and existing range maps have little overlap. In the ocean, two global species distribution datasets, produced by AquaMaps (REF) and the International Union for Conservation of Nature (IUCN) (REF), dominate our understanding of marine species ranges throughout the world's oceans.  Together they represent `r format(nrow(spp_all), big.mark = ',')` species, mostly from AquaMaps, with only `r format(nrow(spp_map_pairs), big.mark = ',')` species overlapping.  Here we examine differences in predicted species ranges between the two datasets, propose mechanistic causes and potential solutions for such differences, and explore the implications of these differences for management and conservation decisions.  We find that IUCN maps often disregard bathymetry for depth-limited species, leading to predictions of species presence at unsuitable depths, and AquaMaps ranges for data-poor species often extrapolate presence far afield from known occurrences.  We illustrate the implications of these differences by repeating two recent applications - the Ocean Health Index (Halpern et al. 2012, 2015) biodiversity goal and a global analysis of gaps in coverage of marine protected areas (Klein et al. 2015) - and find significantly different estimates of the status of global biodiversity and effectiveness of conservation depending on how the two datasets were prioritized. ___Creating a single, highly accurate dataset of global marine species ranges will be difficult. Understanding the implications of dataset differences for conservation planning and decision-making is essential.___
<!--Understanding these issues points toward solutions for providers and users of these datasets.-->

### 2016-03-04 ms_2_intro_res_disc.Rmd
updating refs

One major outcome of this body of science is the various compiled databases of species distribution maps. Two global-scale repositories predict marine species ranges throughout the world's oceans – AquaMaps (Kaschner et al. 2013) and International Union for Conservation of Nature (IUCN) (REF).  These two spatial datasets have been used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, 2015 _both_; Selig et al. 2013 _both_), evaluating global biodiversity patterns (Coll et al. 2010 _AM_; Martin et al. 2014 _both and others_, Pimm 2014 _IUCN, mostly focusing on terrestrial but some marine_), predicting range shifts (Molinos et al. 2015 _AquaMaps_, and setting conservation priorities (Klein et al. 2015 _AquaMaps_). <!-- BSH: Any of these use IUCN?  Would be good to make sure at least 2-3 examples were based on IUCN. ___none of these exclusively use IUCN - other suggestions?___-->

@@

_Four foundational papers relating to AquaMaps (Kaschner et al. 2006, 2011; 2013; Ready et al. 2010) have been cited a total of 379 times._

@@

The IUCN publishes species range maps, bundled by taxonomic groups (REF), based on expert input on spatial boundaries of a given species' "limits of distribution" (REF IUCN 2014) - essentially a refined extent of occurrence, based on observation records and refined by expert understanding of the species' range and habitat preferences.  In contrast, AquaMaps models species distribution based environmental preferences (e.g. temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global raster of probability of occurrence for each species (AquaMaps 2013)

The methodologies behind these datasets imply differences in prediction of species distribution due to errors of commission (falsely _(erroneously? less accusatory, but an annoying word)_ indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps generally include large commission errors, while predicted distribution models such as AquaMaps will likely include fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range <!-- from Rondinini: Omission errors lead to identifying networks of areas for conservation action that are smaller than required and centred on known species occurrences, thus affecting the comprehensiveness, representativeness and efficiency of selected areas. Commission errors lead to selecting areas not relevant to conservation, thus affecting the representativeness and adequacy of reserve network --> (Rondinini, 2006). 
<!-- This study basically predicted some of the problems we’re seeing here and predicts implications; I’d like to meaningfully acknowledge it, but not sure how deeply to get into it.  Later comments ask about whether to include the omission/commission terminology.

@@

To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (where on the map) against the ratio of extent (how much of the map) for each shared species (Fig. 2A).

A general negative linear pattern emerges, <!--BSH: Any stats for this? If so, insert just after this phrase. Otherwise maybe phrase it as 'The general pattern that emerges of a negative linear relationship, where increasing similarity...'--> suggesting that increasing similarity in extent correlates with decreasing distribution alignment. <!--BSH: Could also be said as increasing similarity in alignment leads to decreasing similarity in extent.  The phrasing implies causality, even though you are careful to use ‘correlates’--> AquaMaps tends to extrapolate species ranges beyond known occurrences, such that the marginal range predicted by AquaMaps will fall in different locations than the marginal range predicted using IUCN methodology. For species with dissimilar extents, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar extents, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

@@

* The IUCN Red List instructional materials for mapping marine species suggest that "bathymetry can be used to delineate species' ranges limited by depth in the same way as elevation is used for terrestrial species" (REF https://www.conservationtraining.org/mod/scorm/player.php?a=1412&currentorg=M5L1_ORG&scoid=6492).  Unfortunately, this recommendation appears to be applied inconsistently.  Especially for demersal communities dependent on photosynthesis, bathymetry should be a primary consideration; even a simple cutoff at the photosynthetic limit of 200 meters would minimize a likely source of commission errors.
* The AquaMaps model is a powerful tool for quickly estimating species distribution based on limited information; however, the model's output is only as good as the input data, which for marine species is often sorely limited.  AquaMaps encourages experts to review and comment on the predicted species distribution maps; lessons from the small sample of reviewed maps can be generalized, for example to other species in the same taxonomic group, to further refine the predictions.  This is especially important for data-poor species.
* In addition to environmental preferences and conditions, AquaMaps mapping parameters include area restrictions to help limit over-extrapolation of AquaMaps models.  Currently the area restrictions are denoted by FAO Major Fishing Areas (REF find a reference in AquaMaps somewhere); however, area restrictions based on biogeographical criteria, e.g. Marine Ecoregions of the World (REF), would likely provide better resolution and predictive power, especially for data-poor species.

>>>>

One major outcome of this body of science is the various compiled databases of species distribution maps. Two global-scale repositories predict marine species ranges throughout the world's oceans – AquaMaps (Kaschner et al. 2013) and International Union for Conservation of Nature (IUCN) (REF).  These two spatial datasets have been used for a wide range of purposes, including assessing marine species status (Halpern et al. 2012, 2015 _both_; Selig et al. 2013 _both_), evaluating global biodiversity patterns (Coll et al. 2010 _AM_; Martin et al. 2014 _both and others_, Pimm et al. 2014 _IUCN, mostly focusing on terrestrial but some marine_, Kaschner et al. 2011), predicting range shifts (Molinos et al. 2015 _AquaMaps_, and setting conservation priorities (Klein et al. 2015 _AquaMaps_).

@@

_Three foundational papers relating to AquaMaps (Kaschner et al. 2006, 2011; Ready et al. 2010) have been cited a total of 315 times._

@@

The IUCN publishes species range maps, bundled by taxonomic groups, based on expert input on spatial boundaries of a given species' "limits of distribution" (IUCN 2015) - essentially a refined extent of occurrence, based on observation records and refined by expert understanding of the species' range and habitat preferences.  In contrast, AquaMaps models species distribution based environmental preferences (e.g. temperature, depth, salinity) deduced from occurrence records, published species databases such as FishBase, and expert knowledge.  The AquaMaps model overlays these environmental preferences atop a map of environmental attributes on a 0.5 degree grid, creating a global raster of probability of occurrence for each species (Kaschner et al. 2006, Ready et al. 2010)

The methodologies behind these datasets imply differences in prediction of species distribution due to errors of commission (falsely _(erroneously? less accusatory, but an annoying word)_ indicating species presence) and omission (falsely indicating species absence).  Geographic range data such as IUCN range maps generally include large commission errors, while predicted distribution models such as AquaMaps will likely include fewer commission errors but more omission errors.  Each type of error bears different implications for conservation goals: commission errors can result in prioritizing areas not relevant to conservation goals, while omission errors may result in protected area networks that fail to include important habitat and range <!-- from Rondinini: Omission errors lead to identifying networks of areas for conservation action that are smaller than required and centred on known species occurrences, thus affecting the comprehensiveness, representativeness and efficiency of selected areas. Commission errors lead to selecting areas not relevant to conservation, thus affecting the representativeness and adequacy of reserve network --> (Rondinini et al. 2006). 
<!-- This study basically predicted some of the problems we’re seeing here and predicts implications; I’d like to meaningfully acknowledge it, but not sure how deeply to get into it.  Later comments ask about whether to include the omission/commission terminology.

@@

To explore differences in species distribution and range between the two datasets, we plotted the distribution alignment (where on the map) against the ratio of extent (how much of the map) for each shared species (Fig. 2A). A general negative linear pattern emerges, <!--BSH: Any stats for this? If so, insert just after this phrase. Otherwise maybe phrase it as 'The general pattern that emerges of a negative linear relationship, where increasing similarity...'--> suggesting that increasing similarity in extent correlates with decreasing distribution alignment. <!--BSH: Could also be said as increasing similarity in alignment leads to decreasing similarity in extent.  The phrasing implies causality, even though you are careful to use ‘correlates’--> AquaMaps tends to extrapolate species ranges beyond known occurrences, such that the marginal range predicted by AquaMaps will fall in different locations than the marginal range predicted using IUCN methodology. For species with dissimilar extents, predicted distribution for the smaller range can more easily fall within the generous bounds of the larger range. For species with increasingly similar extents, differences in methodology become more difficult to "hide," and the distribution alignment generally becomes poorer.

@@

* The IUCN Red List instructional materials for mapping marine species suggest that "bathymetry can be used to delineate species' ranges limited by depth in the same way as elevation is used for terrestrial species" (IUCN 2015).  Unfortunately, this recommendation appears to be applied inconsistently.  Especially for demersal communities dependent on photosynthesis, bathymetry should be a primary consideration; even a simple cutoff at the photosynthetic limit of 200 meters would minimize a likely source of commission errors.
* The AquaMaps model is a powerful tool for quickly estimating species distribution based on limited information; however, the model's output is only as good as the input data, which for marine species is often sorely limited.  AquaMaps encourages experts to review and comment on the predicted species distribution maps; lessons from the small sample of reviewed maps could be generalized, for example to other species in the same taxonomic group, to further refine the predictions.  This is especially important for data-poor species.
* In addition to environmental preferences and conditions, AquaMaps mapping parameters include area restrictions to help limit over-extrapolation of AquaMaps models.  Currently area restrictions are denoted by UN Food and Agriculture Organisation (FAO) fisheries reporting areas (Ready et al. 2010); however, area restrictions based on biogeographical criteria, e.g. Marine Ecoregions of the World (Spalding et al. 2007), would likely provide better resolution and predictive power, especially for data-poor species.

### 2016-03-04 ms_5_refs.Rmd
updating refs

__Key AquaMaps publications:__

Kaschner, K., R. Watson, A. W. Trites, D. Pauly (2006). Mapping world-wide distributions of marine mammal species using a relative environmental suitability (RES) model.  Marine Ecology Progress Series 316: 285–310. _check this citation journal name..._ _This outlines the basic RES methodology - AM development_

Kaschner, K., D.P. Tittensor, J. Ready, T Gerrodette and B. Worm (2011). Current and Future Patterns of Global Marine Mammal Biodiversity. PLoS ONE 6(5): e19653. PDF _just what the title says - AM development, presence threshold 60%, also analyzes richness as a function of threshold_

Ready, J., K. Kaschner, A.B. South, P.D Eastwood, T. Rees, J. Rius, E. Agbayani, S. Kullander and R. Froese (2010). Predicting the distributions of marine organisms at the global scale. Ecological Modelling 221(3): 467-478. PDF _Presents AM; assessing AquaMaps against other presence-only species models_

__Papers based on AquaMaps:__

Jones, M.C., S.R. Dyeb, J.K. Pinnegar and W.W.L. Cheung (2012). Modelling commercial fish distributions: Prediction and assessment using different approaches. Ecological Modelling 225(2012): 133-145. PDF _comparison of species distribution models including AquaMaps, Maxent and the Sea Around Us Project_

Coll, M., C. Piroddi, J. Steenbeek, K. Kaschner, F. Ben Rais Lasram et al. (2010). The biodiversity of the Mediterranean Sea: estimates, patterns, and threats. PLoS ONE 5(8): e11842. PDF _used AquaMaps to predict Med biodiversity.  Also: Threshold = 0._

Martin C.S., Fletcher R., Jones M.C., Kaschner K., Sullivan E., Tittensor D.P., Mcowen C., Geffert J.L., van Bochove J.W., Thomas H., Blyth S., Ravillious C., Tolley M., Stanwell-Smith D. (2014). Manual of marine and coastal datasets of biodiversity importance. May 2014 release. Cambridge (UK): UNEP World Conservation Monitoring Centre. 28 pp. (+ 4 annexes totalling 174 pp. and one e-supplement). PDF _report on marine data sets and data gaps etc, incl both IUCN and AM_

-----

Hurlbert 2007 Species richness, hotspots, and the scale dependence of range maps in ecology and conservation. _mostly rasters of range maps?  "The scale dependence of range-map accuracy poses clear limitations on braod-scale ecological analyses and conservation assessments.  ... we provide guidance about the approriate scale of their use_

Jetz 2008 Ecological Correlates and Conservation Implications of Overestimating Species Geographic Ranges _EOO maps are usually highly interpolated and overestimate small-scale occurrence, which may bias research outcomes_

Pimm 2014 The biodiversity of species and their rates of extinction, distribtuion, and protection. _uses range maps to show biodiversity areas; may use IUCN range maps.  Also discusses gaps and possible things that can be done about them._

Rondinini 2006 Tradeoffs of different types of species occurrence data for use in systematic conservation planning _compares point locality, range maps, and distribution models in terms of omission and commission errors; also outlines Extent of Occurrence and Area of Occupancy distinctions._

García Molinos, Jorge, Benjamin S. Halpern, David S. Schoeman, Christopher J. Brown, Wolfgang Kiessling, Pippa J. Moore, John M. Pandolfi, Elvira S. Poloczanska, Anthony J. Richardson, and Michael T. Burrows. “Climate Velocity and the Future Global Redistribution of Marine Biodiversity.” Nature Climate Change advance online publication (August 31, 2015). doi:10.1038/nclimate2769.

Halpern, Benjamin S., Catherine Longo, Darren Hardy, Karen L. McLeod, Jameal F. Samhouri, Steven K. Katona, Kristin Kleisner, et al. “An Index to Assess the Health and Benefits of the Global Ocean.” Nature 488, no. 7413 (August 30, 2012): 615–20. doi:10.1038/nature11397.

Klein, Carissa J., Christopher J. Brown, Benjamin S. Halpern, Daniel B. Segan, Jennifer McGowan, Maria Beger, and James E.M. Watson. “Shortfalls in the Global Protected Area Network at Representing Marine Biodiversity.” Scientific Reports 5 (December 3, 2015): 17539. doi:10.1038/srep17539.

Selig, Elizabeth R., Catherine Longo, Benjamin S. Halpern, Benjamin D. Best, Darren Hardy, Cristiane T. Elfes, Courtney Scarborough, Kristin M. Kleisner, and Steven K. Katona. “Assessing Global Marine Biodiversity Status within a Coupled Socio-Ecological Perspective.” PLoS ONE 8, no. 4 (April 11, 2013): e60284. doi:10.1371/journal.pone.0060284.

* IUCN reference:
    * https://www.conservationtraining.org/course/view.php?id=217&lang=en
    * http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources
    * IUCN Red List - accessed 12/21/2015

* AquaMaps reference, and accessed date?

>>>>

Kaschner K et al. (2015) AquaMaps: Predicted range maps for aquatic species. World wide web electronic publication, www.aquamaps.org, Version 08/2015.

IUCN (2015). The IUCN Red List of Threatened Species. Version 2015-4. <http://www.iucnredlist.org>. Downloaded on 21 December 2015.

Halpern BS et al. "An index to assess the health and benefits of the global ocean." Nature 488.7413 (2012): 615-620. _doi:10.1038/nature11397._

Halpern BS et al. (2015). Patterns and emerging trends in global ocean health. PloS one, 10(3), e0117863.

Klein CJ et al. (2015). Shortfalls in the global protected area network at representing marine biodiversity. Scientific Reports 5: 17539. _doi:10.1038/srep17539._

Selig ER et al. (2013). Assessing global marine biodiversity status within a coupled socio-ecological perspective. PloS one, 8(4), e60284. _doi:10.1371/journal.pone.0060284._

Coll M et al. (2010). The biodiversity of the Mediterranean Sea: estimates, patterns, and threats. PLoS ONE 5(8): e11842. _used AquaMaps to predict Med biodiversity.  Also: Threshold = 0._

*Martin CS et al. (2014). Manual of marine and coastal datasets of biodiversity importance. May 2014 release. Cambridge (UK): UNEP World Conservation Monitoring Centre. 28 pp. (+ 4 annexes totalling 174 pp. and one e-supplement). _review of marine data sets and data gaps etc, incl both IUCN and AM as well as many others_

Pimm SL et al. (2014). The biodiversity of species and their rates of extinction, distribution, and protection. Science, 344(6187), 1246752. _uses range maps to show biodiversity areas; may use IUCN range maps.  Also discusses gaps and possible things that can be done about them._

Kaschner K, Tittensor DP, Ready J, Gerrodette T, Worm B (2011). Current and Future Patterns of Global Marine Mammal Biodiversity. PLoS ONE 6(5): e19653. _just what the title says - AM development, presence threshold 60%, also analyzes richness as a function of threshold_

Molinos JG et al. (2015). Climate velocity and the future global redistribution of marine biodiversity. Nature Climate Change. _doi:10.1038/nclimate2769._

Rondinini C, Wilson KA, Boitani L, Grantham H, & Possingham HP (2006). Tradeoffs of different types of species occurrence data for use in systematic conservation planning. Ecology letters, 9(10), 1136-1145. _compares point locality, range maps, and distribution models in terms of omission and commission errors; also outlines Extent of Occurrence and Area of Occupancy distinctions._

Jetz W, Sekercioglu CH, and Watson JE (2008). Ecological correlates and conservation implications of overestimating species geographic ranges. Conservation Biology, 22(1), 110-119. _EOO maps are usually highly interpolated and overestimate small-scale occurrence, which may bias research outcomes_

Kaschner K, Watson R, Trites AW, Pauly D (2006). Mapping world-wide distributions of marine mammal species using a relative environmental suitability (RES) model.  Marine Ecology Progress Series 316: 285–310. _check this citation journal name..._ _This outlines the basic RES methodology - AM development_

Ready J et al. (2010). Predicting the distributions of marine organisms at the global scale. Ecological Modelling 221(3): 467-478. _Presents AM; assessing AquaMaps against other presence-only species models_

Spalding MD et al. (2007). Marine ecoregions of the world: a bioregionalization of coastal and shelf areas. BioScience, 57(7), 573-583.

### NOT CURRENTLY USED:

Jones MC, Dyeb SR, Pinnegar JK, Cheung WWL (2012). Modelling commercial fish distributions: Prediction and assessment using different approaches. Ecological Modelling 225(2012): 133-145. PDF _comparison of species distribution models including AquaMaps, Maxent and the Sea Around Us Project_

Hurlbert AH, and Jetz W (2007). Species richness, hotspots, and the scale dependence of range maps in ecology and conservation. Proceedings of the National Academy of Sciences, 104(33), 13384-13389. _mostly rasters of range maps?  "The scale dependence of range-map accuracy poses clear limitations on braod-scale ecological analyses and conservation assessments.  ... we provide guidance about the approriate scale of their use_
